---
title: "Final Project Data Analysis for Business - "
output:
  html_document: default
  pdf_document: default
---

```{r message=FALSE, warning=FALSE, include=FALSE}
# Importing libraries

library(ggplot2)
library(dplyr)
library(skimr)
library(readr)
library(sf)
library(usmap)
library(grid)
library(gridExtra)
library(corrplot)
library(caret)
library(ROSE)
library(latex2exp)
library(glmnet)
library(rpart)
library(rpart.plot)
library(randomForest)
library(doParallel)
library(pROC)
library(knitr)
library(xgboost)
library(factoextra)
library(cluster)
library(ggsci)
library(knitr)
library(RColorBrewer)
#

```

# STATBIZ {.tabset .tabset-fade}

Group members:

-   Michele Turco (285251)
-   Giulio Presaghi(287611)
-   Edoardo Brown (281671)
-   Irene Benvenuti (288521)

## TASK 1 {.tabset .tabset-fade}

The dataset under analysis is the "Telecom Churn" dataset, which includes data about a US Telecom company. Our primary goal is to predict customer churn to help the company take proactive steps to retain clients. In a real-world context, understanding this target variable is crucial for businesses aiming to improve customer retention. 'Churn' signifies customers who have ended their relationship with the company, while 'Not churn' indicates those who continue to engage with the company’s services or products. By identifying customers at high risk of leaving, the company can strategically offer promotions and incentives to enhance customer retention and reduce churn rates.

### DATASET DESCRIPTION

#### Importing the dataset

The dataset consists of 3333 observations with 20 variables related to telecom customer attributes, such as state, account length, service usage details, and whether the customer churned or not. In particular, we can see that there are 15 numerical variables and 5 categorical variables.

```{r echo=FALSE, warning=FALSE}
# Importing Dataset
Data = read.csv2("./Dataset/TelecomChurn.csv", 
                 header = T, 
                 sep = ",", 
                 colClasses = "character")
variables = colnames(Data)
categorical_variables = c("State", "International.plan", "Voice.mail.plan", "Area.code")
target_variable = "Churn"
numerical_variables = setdiff(variables, c(categorical_variables, target_variable))
predictors = setdiff(variables, target_variable)

# Ensuring that variables are converted in the correct form
Data[[target_variable]] = as.factor(Data[[target_variable]])
for (var in numerical_variables) {
  Data[[var]] = as.numeric(Data[[var]])
}
for (var in categorical_variables) {
  Data[[var]] = as.factor(Data[[var]])
}

# Showing the results
str(Data)
```

#### Cleaning the dataset

The dataset does not contain any missing value and hence this is not an issue that need to be addressed.

```{r echo=FALSE}

#Renaming missing values with NA notation, and counting how many rows contain missing values, then printing result
na_counts_per_row = rowSums(is.na(Data))
rows_with_na = sum(na_counts_per_row > 0)
cat("Rows with NA before preprocessing:", rows_with_na, "\n")

```

In addition, the dataset does not present any duplicated value and consequently the issue does not need to be addressed.

```{r echo=FALSE}
# Counting duplicates row and printing the result
duplicates = sum(duplicated(Data))
cat("There are", duplicates, "duplicates rows in the Dataset before preprocessing")
```

### EDA {.tabset .tabset-fade}

Exploratory Data Analysis (EDA) is a crucial step in understanding the underlying patterns and relationships within the dataset. Through this analysis, we aim to gain insights that will inform the subsequent steps of data preprocessing and model building.
We will consider categorical variables (excluding the state variable because of its nature) and numerical variables, for which we computed a correlation matrix.

#### Target variable distribution

In this section we consider our target variable "Churn" and its distribution. To do this, we will plot both a histogram to visualize the number of observations for each class and a pie chart to better understand their proportions in the dataset. 

```{r include=FALSE}

# Computing churn count and proportion
churn_distribution = Data %>%
  count(Churn, name = "Count")
churn_distribution$proportion = churn_distribution$Count/sum(churn_distribution$Count)

# Creating histogram with ggplot library
histogram_plot = ggplot(Data, aes(x = Churn, fill = Churn)) + 
  geom_bar(color = "black", alpha = 0.7) +
  geom_text(stat='count', aes(label=after_stat(count)), vjust= 2) + 
  ggtitle("Churn Count") + 
  scale_fill_manual(values = c("red", "blue")) + 
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 10),
    legend.position = "none",
  )

# Creating pie histogram plot with ggplot library
piechart_plot = ggplot(data = churn_distribution, aes(x = "", y = proportion, fill = Churn)) + 
  geom_bar(stat = "identity", color = "black", alpha = 0.7) +
  coord_polar("y") +
  theme_minimal() + 
  geom_text(aes(label = paste0(round(proportion * 100, 0), "%")), 
            position = position_stack(vjust = 0.5)) + 
  labs(x = NULL, y = NULL, fill = NULL, 
       title = paste("Churn Proportion")) + 
  scale_fill_manual(values = c("red", "blue", "green")) +
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 12),
        panel.grid = element_blank())

```

```{r echo=FALSE}

# Representing the plots

grid.arrange(
  arrangeGrob(histogram_plot, piechart_plot, nrow = 1, ncol = 2)
)

```

The histogram on the left shows that there are 2850 customers who did not churn (False) and 483 customers who did churn (True). The pie chart on the right indicates that 86% of customers did not churn, while 14% did churn. This highlights an significance imbalance in the dataset, with an higher number of customers not churning compared to those who do.

#### Categorical variables

In this section we analyze the categorical variables in our dataset. In order to get more insightful results in the analysis of the distribution of categorical variables, we firstly plot the proportion of each class in a pie chart (as we did before with our target variable).

```{r include=FALSE}
# Code used in order to generate the plots

plot_list = list()
plot_list_relationship = list()

# Loop through the categorical variables to create individual plots
for (variable in categorical_variables) {
  if (variable == "State") {
    next
  }
  
  # Creating histograms
  plot_relationship = ggplot(Data, aes_string(x = variable, fill = "Churn")) + 
    geom_bar(position = "dodge", color = "black", alpha = 0.7) + 
    scale_fill_manual(values = c("blue", "red")) + 
    geom_text(stat = 'count', aes(label = after_stat(count)), position = position_dodge(width = 1), vjust = -0.4) +
    xlab(variable) +
    ylim(0, 2700) +
    ylab("Count") + 
    theme_minimal() + 
    theme(
      plot.title = element_text(hjust = 0.5, size = 16),
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      axis.text.x = element_text(size = 10),
      axis.text.y = element_text(size = 10),
      legend.title = element_text(size = 14),
      legend.text = element_text(size = 12),
      legend.position = "none"
    )
  
  # Storing the legend separately for histograms (we want to represent only one in the final plot)
  get_legend <- function(myplot) {
    tmp <- ggplot_gtable(ggplot_build(myplot))
    leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
    legend <- tmp$grobs[[leg]]
    return(legend)
  }
  legend = get_legend(ggplot(Data, aes_string(x = categorical_variables[1], fill = "Churn")) + 
                         geom_bar(position = "dodge", color = "black", alpha = 0.7) + 
                         scale_fill_manual(values = c("blue", "red")) + 
                         theme_minimal() + 
                         theme(legend.title = element_text(size = 14),
                               legend.text = element_text(size = 12)))
  
  # Computing proportions for categorical variables
  count_df = Data %>%
    count(!!sym(variable), name = "Count")
  count_df$proportion = count_df$Count/sum(count_df$Count)
 
  # Creating pie charts
  plot = ggplot(data = count_df, aes(x = "", y = proportion, fill = !!sym(variable))) + 
    geom_bar(width = 1, stat = "identity", color = "black", alpha = 0.7) + 
    theme_classic() + 
    coord_polar("y") +
    geom_text(aes(label = paste0(round(proportion * 100, 0), "%")), 
              position = position_stack(vjust = 0.7)) + 
    labs(x = NULL, y = NULL, fill = NULL, 
         title = paste("Distribution of", variable)) + 
    guides(fill = guide_legend(reverse = TRUE)) + 
    scale_fill_manual(values = c("red", "blue", "green")) + 
    theme(axis.line = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank(),
          plot.title = element_text(hjust = 0.5))
  

  # Adding the plots to the list
  plot_list_relationship[[variable]] = plot_relationship
  plot_list[[variable]] = plot
}

```

```{r echo=FALSE}
# Display the first plot
grid.arrange(
  arrangeGrob(grobs = plot_list, nrow = 2, ncol = 2)
)
```

The pie chart shows that 90% of the customers do not have an international plan, while only 10% have opted for one. This indicates that the majority of customers do not use international calling services. In addition, 72% of the customers do not have a voice mail plan, whereas 28% have subscribed to this service. This suggests that a significant portion of the customer base does not utilize voice mail services. The pie chart shows also the distribution of customers across three area codes: 408, 415, and 510. The distribution is relatively balanced, with 50% of customers having area code 415, and the remaining 50% evenly split between area codes 408 and 510 (25% each). This indicates that the customer base is fairly evenly distributed across these three regions.

Furthermore, we wanted to highlight the distribution of customer churn across the three categorical variables with an histogram in which the bars represent the counts of customers who churned (True) and did not churn (False) for each category.

```{r echo=FALSE}
grid.arrange(
  arrangeGrob(grobs = plot_list_relationship, nrow = 1, ncol = 3),
  top = textGrob("Churn Distribution Across Categorical Variables", gp = gpar(fontsize = 10, fontface = "bold"), just = "center"),
  right = legend,
  left = "Count"
)
```

The majority of customers without an international plan (2664) did not churn, whereas 346 customers did churn. For customers with an international plan, there are 186 non-churners compared to 137 churners.

Among customers without a voice mail plan, 2008 did not churn, while 403 churned. For customers with a voice mail plan, 842 did not churn and 80 churned.

The churn distribution is somewhat balanced across area codes, with 408 and 510 having slightly higher churn rates (122 and 125 churners respectively) compared to area code 415 (236 churners). The number of non-churners is highest in area code 415 (1419) compared to 408 (716) and 510 (715).

What is particularly interesting is that the proportion of churn is higher among customers with an international plan compared to those without. In fact, almost half of the customers who were subscribed to the International plan have churned, suggesting that this factor may heavily influence customer decisions. However, to draw any definitive conclusions, we need to perform a more in-depth analysis.

##### Chi-Square Test with International Plan and Churn

To statistically interpret the hypothesis stated above, we performed a Chi-Square Test on the two variables. The Chi-Square test is a statistical method used to determine if there is a significant association between two categorical variables. It compares the observed frequencies in each category to the expected frequencies under the null hypothesis, which posits that the variables are independent.

-   H0 (Null Hypothesis): There is no association between the two categorical variables; they are independent.

-   H1 (Alternative Hypothesis): There is an association between the two categorical variables; they are not independent.

```{r echo=FALSE}
# Create a contingency table and perform the chi-squared test. Then we report the results
contingency_table = table(Data$International.plan, Data$Churn)
chi_squared_test = chisq.test(contingency_table)
print("Chi Square Test on International Plan and Churn")
print(chi_squared_test)
```

X-squared = 222.57 is the test statistic value, which measures the discrepancy between the observed and expected frequencies. The extremely small p-value (\< 2.2e-16) indicates that the probability of observing such a discrepancy by random chance is very low. Since the p-value is much smaller than the conventional significance level (e.g., 0.05), the null hypothesis is rejected, suggesting that there is a statistically significant association between the two categorical variables in the dataset. In other words, the data provides strong evidence that the variables are not independent and that there is a significant relationship between them.

##### State variable

Due to its particularly high number of levels (51), we decided to analyze this variable differently from other categorical variables. A "classic" plot would not have been as insightful. Therefore, we plotted the churn rate across USA states directly on a map, (after ensuring rthat observations were balanced between states). This approach provides a clearer understanding of whether the variable distribution varies by state.

```{r echo=FALSE}
Data$ChurnNumeric = ifelse(Data$Churn == "True", 1, 0)

churn_rate_states = Data %>%
  group_by(State) %>%
  summarize(ChurnRate = mean(ChurnNumeric))

states = statepop
names(states)[names(states) == "abbr"] <- "State"
churn_rate_states = merge(states, churn_rate_states, by = "State", all.x = TRUE)

plot_usmap(data = churn_rate_states, values = "ChurnRate", labels = TRUE) +
  scale_fill_gradient(low = "lightblue",
                      high = "red",
                      name = NULL) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.direction = "horizontal",
        legend.key.width = unit(0.8, "in"),
        plot.title = element_text(hjust = 0.5)) +
  ggtitle("Churn Rate Across US States")

# Dropping variables since it is now useless
Data$ChurnNumeric = NULL
```

The plot indicates that the churn rate distribution does vary significantly across states. For example, California (CA) and Texas (TX) exhibit higher churn rates, indicated by the deep red shading (25%, as shown in the legend), suggesting that a substantial proportion of customers in these states are leaving the service. In contrast, states like Alaska (AK) and Iowa (IA), shaded in blue, have much lower churn rates (\<10%), indicating better customer retention.

Reference -> https://youtu.be/Hi3fXGRBCMA?si=LfPkHpsmNk6cQ_hZ

##### Chi squared test on State and Churn

We decided to perform the Chi-Square test in order to have a more accurate insight.

```{r echo=FALSE, warning=FALSE}
# Create a contingency table and perform the chi-squared test. Then we report the results
contingency_table = table(Data$State, Data$Churn)
chi_squared_test = chisq.test(contingency_table)
print("Chi Square Test on State and Churn")
print(chi_squared_test)
```

Since the p-value is less than 0.05 also in this case, we reject the null hypothesis (H0) that the two categorical variables are independent. This indicates a significant association between the variables. However, while both tests reject the null hypothesis, the first test shows a more pronounced discrepancy between observed and expected frequencies with a simpler model, while the second test shows a significant association in a more complex scenario, shown by the degrees of freedom df = 50 (higher number of categories for the state variable).

#### Numerical variables

In this section, we perform Exploratory Data Analysis (EDA) on numerical variables to understand their distributions, identify patterns, and detect any anomalies. This analysis will provide key insights and inform subsequent data preprocessing and modeling steps.

```{r include=FALSE, warning=FALSE}

# Code to generate the plots
Numerical_Data = Data[numerical_variables]
means_vec = apply(X = Numerical_Data, MARGIN = 2, FUN = mean)
median_vec = apply(X = Numerical_Data, MARGIN = 2, FUN = median)
sd_vec = apply(X = Numerical_Data, MARGIN = 2, FUN = sd)

plot_list_numerical_relationship = list()
plot_list_numerical = list()

# Loop through each numerical variable to create individual histograms
for (variable in numerical_variables) {
  # Setting appropriate bin width
  binwidth = ceiling(max(Data[[variable]], na.rm = TRUE)/10)
  # Removing points from the variables names
  name = gsub("\\.", " ", variable)
  
  # Creating histograms
  plot_numerical = ggplot(Data, aes_string(x = variable)) + 
    geom_histogram(binwidth = binwidth, color = "black", fill = "grey", alpha = 0.7) + 
    geom_vline(aes_string(xintercept = means_vec[variable]), color = "blue", linetype = "dashed", size = 1) + 
    geom_vline(aes_string(xintercept = median_vec[variable]), color = "red", linetype = "dashed", size = 1) + 
    labs(title = paste("Histogram of", variable), x = name, y = "Frequency") +
    theme_minimal() + 
    theme(
      plot.title = element_blank(),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_blank(),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12)
    )
  
  # Creating 
  plot_numerical_relationship = ggplot(Data, aes_string(x = variable, fill = "Churn")) +
    geom_histogram(position = "identity", alpha = 0.7, binwidth = binwidth) +
    labs(title = variable, x = name) +
    theme_minimal() +
    theme(
      plot.title = element_blank(),
      axis.title.x = element_text(size = 12),
      axis.title.y = element_blank(),
      axis.text.x = element_text(size = 10),
      axis.text.y = element_text(size = 10)
    ) +
    scale_fill_manual(name = "Churn", labels = c("No", "Yes"), values = c("blue", "red"))
  
  # Storing legend
  legend = get_legend(ggplot(Data, aes_string(x = variable, fill = "Churn")) +
                        geom_histogram(position = "identity", alpha = 0.7, binwidth = binwidth) +
                        labs(title = variable) +
                        theme_minimal() +
                        theme(legend.title = element_text(size = 14),
                              legend.text = element_text(size = 12),
                              legend.direction = "horizontal") +
                        scale_fill_manual(name = "Churn", labels = c("No", "Yes"), values = c("blue", "red"))
                      ) 
  
  # Add the plot to the list
  plot_list_numerical[[variable]] = plot_numerical
  plot_list_numerical_relationship[[variable]] = plot_numerical_relationship
}

```


```{r fig.width=10, fig.height=10, echo=FALSE, warning=FALSE, include=FALSE}

grid.arrange(
  arrangeGrob(grobs = plot_list_numerical, nrow = 5, ncol = 3),
  top = textGrob("Numerical Variables Distribution", gp = gpar(fontsize = 20, fontface = "bold"))
)

```


```{r fig.width=10, fig.height=10, echo=FALSE, warning=FALSE}

grid.arrange(
  arrangeGrob(grobs = plot_list_numerical_relationship, nrow = 5, ncol = 3),
  top = textGrob("Numerical Variables Distribution", gp = gpar(fontsize = 20, fontface = "bold")),
  bottom = legend,
  left = textGrob("Count", rot = 90, gp = gpar(fontsize = 14))
)

```
The histograms reveal that several variables, such as Account length, Total day minutes, and Total night charge, exhibit relatively symmetric distributions approximating normality. In contrast, variables like Number vmail messages and Customer service calls display right skewness, where most values are low with a few higher values. Notably, Number vmail messages and Total intl calls have significant zero values, suggesting under utilization of these services. The red dashed lines represent the means, indicating the central tendency for each variable.

Across most variables, the proportion of customers who churn (red) is relatively small compared to those who do not churn (blue), consistent with the overall churn rate in the dataset. For Customer service calls, the churn proportion noticeably increases with higher values, suggesting that frequent contact with customer service may indicate dissatisfaction leading to churn. In Total day minutes, customers with very high usage show a slightly higher churn rate, potentially due to issues related to usage limits or service quality.

Consequently, we decided to better explore the relationships between the target variable and these two categorical variables.

##### ANOVA test on Total Day Minutes and Churn

The ANOVA (Analysis of Variance) test between a categorical and a continuous variable is a statistical method used to determine if there are significant differences in the means of the continuous variable across the levels of the categorical variable. It tests the null hypothesis that the means are equal across all groups against the alternative hypothesis that at least one group mean is different.

In the first place, we better represent the data with boxplots.

```{r echo=FALSE}

ggplot(Data, aes(x = Churn, y = `Total.day.minutes`, fill = Churn)) + 
  geom_boxplot() + 
  ggtitle("Total Day Minutes by Churn Status") + 
  xlab("Churn") + 
  ylab("Total Day Minutes") +
  scale_fill_manual(values = c("red", "blue")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
  )

```

Churners (blue) have a higher median Total Day Minutes compared to non-churners (red). The spread of Total Day Minutes is similar for both groups, but non-churners have more outliers with lower usage. This confirms that higher Total Day Minutes among churners suggest that heavy usage might be linked to a higher likelihood of churn, while the inactive users (outliers) do not churn at all.

```{r echo=FALSE, warning=FALSE}


# Perform an ANOVA to test the hypothesis H0 that the two means (of the groups),
# are the same, suggesting that having more minutes is not a valid indicator for Churn.

anova_total_minutes = aov(Total.day.minutes ~ Churn, data = Data)
print("ANOVA Test on Total day minutes and Churn")
summary(anova_total_minutes)
```

The ANOVA test shows a significant effect of Churn on the continuous variable analyzed, with a very high F value (146.4) and a p-value \< 2e-16. This indicates that the means of Total day minutes differ significantly between the churn and non-churn groups.

##### ANOVA test on Customer Service Calls and Churn

We repeat the same test for Customer Service Calls

```{r echo=FALSE, warning=FALSE}

anova_service_calls = aov(Customer.service.calls ~ Churn, data = Data)
print("ANOVA Test on Customer Service Calls and Churn")
summary(anova_service_calls)

```

In this case, the ANOVA results show a highly significant effect of the churn status on the continuous variable, with a very low p-value (\< 2e-16). This indicates a significant difference in the means of the continuous variable between customers who churn and those who do not. The high F value (151.8) suggests that the churn status explains a substantial portion of the variability in the continuous variable, more so than indicated by the previous ANOVA results.

#### Correlation Analysis

In this section, we measure the correlation between numerical variables. Correlation is a statistical measure that describes the extent to which two variables change together. It is particularly relevant since collinearity, or multicollinearity, occurs when two or more predictor variables in a dataset are highly correlated. This means that these variables share similar information and move together in a predictable way. We will deal later with this aspect.

```{r fig.width=10, echo=FALSE}

### COLLINEARITY

cor_matrix = cor(Data[numerical_variables], use="complete.obs")
corrplot(cor_matrix, method = "color", tl.srt = 45, tl.col = "black",
         addCoef.col = "black", 
         number.cex = 0.7,
         addgrid.col = "grey",
         tl.cex = 0.8,
         col = colorRampPalette(c("red", "white", "blue"))(200))

```

This correlation matrix shows the relationships between various numerical variables in the dataset. It reveals that no variables are strongly correlated with each other. However, there are some variables that are essentially the same quantity, such as Total day minutes and Total day charge, Total eve minutes and Total eve charge, and Total night minutes and Total night charge, as indicated by their perfect or near-perfect correlations (correlation coefficients close to 1). This is because the total charge for each type is computed by multiplying the total minutes by a constant (or nearly constant) factor. For instance, in the case of day charge, the factor is 0.17. Hence, these pairs of variables provide the same information.

### PREPROCESSING {.tabset .tabset-fade}

#### Dealing with collinearity

As stated during our EDA, considering collinearity and correlations is crucial in classification problems because highly correlated variables can lead to redundancy and affect the model's performance. When predictors are highly correlated, they may provide overlapping information, which can complicate the model's ability to learn effectively and can also inflate the importance of certain features. By identifying and addressing these correlations, we can simplify the model, improve its interpretability, and enhance its predictive performance. Consequently, we proceed in removing variables that are highly correlated to others.


```{r echo=FALSE}

# We have seen in our EDA that there are more features with a really high correlation
# and hence we have to drop them.

cor_matrix[!lower.tri(cor_matrix)] = 0

# Find the pairs with correlation greater than the threshold
threshold = 0.8
high_corr_pairs = which(abs(cor_matrix) > threshold, arr.ind = TRUE)

# Create a data frame with the results
high_corr_df = data.frame(
  Variable1 = rownames(cor_matrix)[high_corr_pairs[, 1]],
  Variable2 = colnames(cor_matrix)[high_corr_pairs[, 2]],
  Correlation = cor_matrix[high_corr_pairs]
)

variables_to_drop = high_corr_df$Variable1

for (variable in variables_to_drop) {
  Data[[variable]] = NULL
  cat("Dropped", variable)
  cat("\n")
  numerical_variables = setdiff(numerical_variables, variable)
  variables = setdiff(variables, variable)
  predictors = setdiff(predictors, variable)
}

```

#### Dealing with categorical variable with too many levels

Having a categorical variable with too many levels in the dataset can increase model complexity, lead to overfitting, reduce computational efficiency, and make the model harder to interpret. In our case, the variable State has too many levels and hence needs to be preprocessed to reduce its dimension. After some research, we found that since 1950, the United States Census Bureau (a principal agency of the U.S. Federal Statistical System, responsible for producing data about the American people) defines four statistical regions: Northeast, West, South, and Midwest. This division is widely used for data collection and analysis and fits our case particularly well as it reduces the levels of the variable from 51 to 4, a significant improvement.


Reference -> https://en.wikipedia.org/wiki/List_of_regions_of_the_United_States


```{r echo=FALSE}
# Create a list to store regions and their states
dic = list()
dic$Northeast = c("CT", "ME", "MA", "NH", "NJ", "NY", "PA", "RI", "VT")
dic$Midwest = c("IL", "IN", "IA", "KS", "MI", "MN", "MO", "NE", "ND", "OH", "SD", "WI")
dic$South = c("AL", "AR", "DE", "DC", "FL", "GA", "KY", "LA", "MD", "MS", "NC", "OK", "SC", "TN", "TX", "VA", "WV")
dic$West = c("AK", "AZ", "CA", "CO", "HI", "ID", "MT", "NV", "NM", "OR", "UT", "WA", "WY")


# Function to find the region for a state
find_region <- function(state) {
  for (region in names(dic)) {
    if (state %in% dic[[region]]) {
      return(region)
    }
  }
  return(NA)
}

# Apply the function to each state in the data frame to add the region column
Data$Region = sapply(Data$State, find_region)
cat("Distinct values for the Region variable:", unique(Data$Region))


# Converting region in a factor and dropping (but storing) the State variable
Data$Region = as.factor(Data$Region)
states = Data$State
Data$State = NULL
variables = setdiff(variables, "State")
categorical_variables = setdiff(categorical_variables, "State")
predictors = setdiff(predictors, "State")
variables = c(variables, "Region")
categorical_variables = c(categorical_variables, "Region")
predictors = c(predictors, "Region")

```

### SPLITTING AND SCALING DATASET

In the first place, we decided to split the dataset into a training set (75% of the data) and a test set (25% of the data). We examined the distribution of the target variable Churn for the original, training, and test sets, displaying the counts and proportions of churned vs. non-churned customers.

```{r echo=FALSE}
### NORMAL SPLIT

set.seed(1)

id_train = sample(1:nrow(Data), size = 0.75*nrow(Data), replace = F)
train_data = Data[id_train,]
test_data = Data[-id_train,]

# Response variable distribution in the original data
cat("Distribution of the target variable in the original set:\n")
cat("Counts:")
print(table(Data$Churn))
cat("Proportions:")
print(prop.table(table(Data$Churn)))
cat("\n")

# Response variable distribution in the train test
cat("Distribution of the target variable in the train set:\n")
cat("Counts:")
print(table(train_data$Churn))
cat("Proportions:")
print(prop.table(table(train_data$Churn)))
cat("\n")

# Response variable distribution in the test set
cat("Distribution of the target variable in the validation set:\n")
cat("Counts:")
print(table(test_data$Churn))
cat("Proportions:")
print(prop.table(table(test_data$Churn)))

```

To address the unbalanced dataset, we created a training set using oversampling. Oversampling is important as it provides the model with enough examples of the minority class, helping it learn the characteristics of both classes more accurately. This improves the model's ability to predict minority class instances, which is crucial in our analysis since predicting customer churn is a key objective for Telecom.

We also attempted undersampling, as suggested in the R reference script of the project. However, the results were quite poor due to the reduced number of data points available for training, which limited the model's learning capability. Consequently, we did not report our findings using this technique.

```{r echo=FALSE}

### OVERSAMPLING

set.seed(1)

# Calculate the total number of samples needed for balanced oversampling
# We want each class to have max_class_count samples
target_N <- 2 * max(table(train_data$Churn))

# Perform oversampling
train_data_oversample <- ovun.sample(Churn ~ ., data = train_data, method = "over", N = target_N)$data

# Response variable distribution in the train set with over sampling
cat("Distribution of the target variable in the oversampled test set:\n")
cat("Counts:")
print(table(train_data_oversample$Churn))
cat("Proportions:")
print(prop.table(table(train_data_oversample$Churn)))

```

Afterwards, we scaled the numerical variables in the training and oversampled datasets using their respective means and standard deviations. Scaling ensures that all numerical features contribute equally to the model’s learning process, preventing features with larger ranges from dominating the model’s behavior. We also scaled the test set using the means and standard deviations from the original training set, ensuring consistency in data preprocessing.

```{r include=FALSE, warning=FALSE}

### SCALING

cols_to_scale = numerical_variables

# Normal data
train_mean = apply(train_data[, cols_to_scale], MARGIN = 2, FUN = mean)
train_sd = apply(train_data[, cols_to_scale], MARGIN = 2, FUN = sd)
train_data_scaled = train_data
train_data_scaled[, cols_to_scale] = scale(train_data[, cols_to_scale], 
                                                      center = train_mean, 
                                                      scale = train_sd)

# Oversample data
train_oversample_mean = apply(train_data_oversample[, cols_to_scale], MARGIN = 2, FUN = mean)
train_overssample_sd = apply(train_data_oversample[, cols_to_scale], MARGIN = 2, FUN = sd)
train_data_scaled_oversample = train_data_oversample
train_data_scaled_oversample[, cols_to_scale] = scale(train_data_oversample[, cols_to_scale], 
                                                      center = train_oversample_mean, 
                                                      scale = train_overssample_sd)


# Scale validation data using training data's parameters
test_data_scaled = test_data
test_data_scaled[, cols_to_scale] = scale(test_data_scaled[, cols_to_scale], 
                                          center = train_mean, 
                                          scale = train_sd)

test_data_scaled_oversample = test_data
test_data_scaled_oversample[, cols_to_scale] = scale(test_data_scaled_oversample[, cols_to_scale], 
                                                     center = train_oversample_mean, 
                                                     scale = train_overssample_sd)

```

### LOW DIMENSIONAL MODEL ANALYSIS {.tabset .tabset-fade}

Before proceeding to the full model aimed at providing the best result, we focus on some lower-dimensional models by considering only a few variables. This approach helps investigate interesting relationships within the data, offering clearer insights and interpretations

#### Effects of Total day minutes and International Plan on Customer Churn

As a first analysis, we fit a logistic regression model to evaluate the effects of Total day minutes and International Plan on Customer Churn, considering the unbalanced dataset.

```{r}

set.seed(1)
model_international_minutes = glm(Churn ~ Total.day.minutes * International.plan, 
                           family = binomial(link = "logit"), 
                           data = test_data)
# View model summary
summary(model_international_minutes)

```
- The Intercept (-4.937732) is the log-odds of churn when Total.day.minutes is 0 and there is no international plan. It is highly significant with a very low p-value (< 2e-16), indicating a strong baseline effect. The result is consistent with the dataset highly unbalanced dataset on which the model was trained, since it indicates a general propensity of not to churn.
- The positive coefficient for Total day minutes suggests that higher Total.day.minutes is associated with an increased likelihood of churn. This effect is statistically significant (p < 2e-16), confirming the result of the ANOVA test in our EDA.
- The coefficient for the International plan is 4.231389. This indicates that having an international plan significantly increases the odds of churn also because of its high significance level (p = 5.20e-05). This is something that confirms the hypothesis tested during our EDA.
- The interaction term is negative and significant (p = 0.0171), even if not as the other terms, indicating that the effect of Total.day.minutes on churn is different for those with an international plan. Specifically, for customers with an international plan, the increase in churn likelihood per minute of day usage is less than for those without an international plan.

We proceed by plotting the computed probabilities.
```{r echo=FALSE, fig.width=10}

# Computing and storing predictive probabilities
Data$predicted_probabilities = predict(model_international_minutes, newdata = Data, type="response")

#Plotting computed probabilities
ggplot(Data, aes(x = Total.day.minutes, y = predicted_probabilities, color = International.plan)) + 
  geom_line(size = 1.2) + 
  labs(title = "Probability of Churn by Total Day Minutes and International Plan", 
       y = "Probability of Churn", x = "Total Day Minutes") +
  scale_color_manual(values = c("red", "blue")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    axis.title.x = element_text(size = 14, face = "bold"),  
    axis.title.y = element_text(size = 14, face = "bold"),  
    axis.text.x = element_text(size = 12, color = "black"),  
    axis.text.y = element_text(size = 12, color = "black"),  
    panel.grid.major = element_line(color = "gray", linewidth = 0.5),  
    panel.grid.minor = element_line(color = "lightgray", linewidth = 0.25),  
    panel.background = element_rect(fill = "white") 
  )

# We drop the column of predicted probabilities since it is now useless
Data$predicted_probabilities = NULL

```

The plot illustrates the probability of churn as a function of Total Day Minutes and the presence of an International Plan, based on an unbalanced training dataset. The red curve represents customers without an international plan, and the blue curve represents those with an international plan. The plot confirms our interpretation of the coefficients: even with an International Plan and with a high value of total day minutes, the predicted probability of churn is lower than 0.5. This suggests that such an unbalanced dataset may negatively affect the predictive performance of the model.

Specifically, for customers without an international plan (red line), the probability of churn increases sharply as total day minutes increase, reflecting a higher sensitivity to usage. In contrast, for customers with an international plan (blue line), the probability of churn remains relatively stable and lower, even as total day minutes increase. This aligns with the negative interaction term in the model, indicating that the increase in churn likelihood per minute of day usage is less for customers with an international plan compared to those without.

As a consequence, to address the imbalance and improve model accuracy, it is essential to consider training the model on a balanced dataset, ensuring that the model captures the true effect of predictors on churn, providing more reliable predictions.

```{r }

set.seed(1)
model_international_minutes_oversample = glm(Churn ~ Total.day.minutes * International.plan, 
                                              family = binomial(link = "logit"), 
                                              data = train_data_oversample)
```

```{r include=FALSE}

# View model summary
summary(model_international_minutes_oversample)

```

```{r echo=FALSE, fig.width=10}

# Computing and storing predictive probabilities
Data$predicted_probabilities = predict(model_international_minutes_oversample, 
                                        newdata = Data, 
                                        type="response")

#Plotting computed probabilities
ggplot(Data, aes(x = Total.day.minutes, y = predicted_probabilities, color = International.plan)) + 
  geom_line(size = 1.2) + 
  labs(title = "Probability of Churn by Total Day Minutes and International Plan (Balanced)", 
       y = "Probability of Churn", x = "Total Day Minutes") +
  scale_color_manual(values = c("red", "blue")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    axis.title.x = element_text(size = 14, face = "bold"),  
    axis.title.y = element_text(size = 14, face = "bold"),  
    axis.text.x = element_text(size = 12, color = "black"),  
    axis.text.y = element_text(size = 12, color = "black"),  
    panel.grid.major = element_line(color = "gray", size = 0.5),  
    panel.grid.minor = element_line(color = "lightgray", size = 0.25),  
    panel.background = element_rect(fill = "white") 
  )

# We drop the column of predicted probabilities since it is now useless
train_data_oversample$predicted_probabilities = NULL

```

As we can see, the balanced dataset corrects the underestimation of churn probabilities observed in the unbalanced dataset, providing a clearer picture of the churn dynamics.
For customers without an international plan, the probability of churn now exceeds 0.5 much sooner, even at moderate levels of Total Day Minutes, reflecting a more accurate assessment of churn risk.
For customers with an international plan, the probability of churn is consistently higher compared to the unbalanced dataset, but still lower than the no-plan group, aligning with the interaction effect observed in the model.


#### Effects of Customer Service Calls on Churn

his analysis aims to evaluate the impact of 'Customer Service Calls' on customer churn using logistic regression models. Given that probabilities are underestimated with an imbalanced dataset, we will only consider the oversampled dataset for this evaluation. The analysis of the imbalanced dataset is still available in the reference R script and revealed the same results as before.

Hence, we build the model using only customer service calls and we plot the computed probabilities.

```{r}

set.seed(1)
model_customer_service_calls_oversample = glm(Churn ~ Customer.service.calls, 
                                             family = binomial(link = "logit"), 
                                             data = train_data_oversample)

```

```{r echo=FALSE}

# Computing and storing predictive probabilities
Data$predicted_probabilities = predict(model_customer_service_calls_oversample, 
                                        newdata = Data, 
                                        type="response")

#Plotting computed probabilities
ggplot(Data, aes(x = Customer.service.calls, y = predicted_probabilities)) + 
  geom_line(size = 1.2, color = "blue") + 
  labs(title = "Probability of Churn by Customer Service Calls (Balanced)",
       y = "Probability of Churn", x = "Customer Service Calls") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),  
    axis.title.x = element_text(size = 14, face = "bold"),  
    axis.title.y = element_text(size = 14, face = "bold"),  
    axis.text.x = element_text(size = 12, color = "black"),  
    axis.text.y = element_text(size = 12, color = "black"),  
    panel.grid.major = element_line(color = "gray", size = 0.5),  
    panel.grid.minor = element_line(color = "lightgray", size = 0.25),  
    panel.background = element_rect(fill = "white") 
  ) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.1))

# We drop the column of predicted probabilities since it is now useless
train_data_oversample$predicted_probabilities = NULL

```


Again, the model confirms our findings during the EDA process. The plot demonstrates a strong positive relationship between the number of customer service calls and the probability of churn, highlighting the importance of effective customer service. The relationship is however quite trivial: frequent interactions with customer service often indicate that the customer is experiencing recurring issues or dissatisfaction with the service. The more frequently a customer contacts customer service, the more likely they are to be facing unresolved problems, leading to frustration and a higher likelihood of churn.

#### Classification tree

To provide a more concrete interpretation of our findings, we decided to implement a small but highly interpretable classification tree. This model is ideal for performing a clear analysis as it allows us to easily visualize and understand the decision paths and key factors influencing customer churn. The simplicity and transparency of a classification tree make it an excellent choice for communicating insights and actionable recommendations based on our data.

```{r}

set.seed(1)
simple_tree_oversample = rpart(Churn ~ Total.day.minutes + International.plan + Customer.service.calls, 
                               data = train_data_oversample, 
                               method = "class")

```

```{r echo=FALSE}

# Plotting tree
rpart.plot(simple_tree_oversample, main = "Classification Tree")

```

The classification tree shows that customers with high total day minutes (>= 223) (32% in the balanced dataset) have a 75% probability of churn. For customers with fewer total day minutes (< 223), churn probability is influenced by customer service calls and the presence of an international plan. Those with less than 4 customer service calls and no international plan (44% of the observations) have a low churn probability (15%), whereas having an international plan in this group (10%) increases the churn probability to 75%. Additionally, customers with 4 or more customer service calls have the highest churn probability (87%), confirming again the relevance of this factor on customer retention.


### BEST MODEL SELECTION {.tabset .tabset-fade}

```{r include=FALSE}

#### Function to compare performances

get.metrics<- function(conf.mat) {
  true.positives <- conf.mat[2,2]
  true.negatives <- conf.mat[1,1]
  false.positives <- conf.mat[1,2]
  false.negatives <- conf.mat[2,1]
  num.observations <- true.positives + true.negatives + false.positives + false.negatives
  
  accuracy <- (true.positives + true.negatives) / num.observations
  precision <- (true.positives) / (true.positives + false.positives)
  recall <- true.positives / (true.positives + false.negatives)
  f1 <- 2 * ((precision * recall) / (precision + recall))
  
  metrics <- data.frame(t(c(accuracy, precision, recall, f1)))
  columns <- c("Accuracy", "Precision", "Recall", "F1")
  colnames(metrics) <- columns
  
  return(metrics)
}

```

```{r include=FALSE}
# Function to plot metrics
plot_table = function(df) {
  
  table_theme = ttheme_default(
    core = list(
      bg_params = list(fill = c(rep(c("white"), each = nrow(df)), NA), col = "black"),
      fg_params = list(fontface = 1, fontsize = 16)
    ),
    colhead = list(
      bg_params = list(fill = "lightblue", col = "black"),
      fg_params = list(fontface = 2, fontsize = 18)
    ),
    rowhead = list(
      bg_params = list(fill = "lightblue", col = "black"),
      fg_params = list(fontface = 2, fontsize = 18)
    )
  )
  table_plot = tableGrob(df, theme = table_theme)
  return(table_plot)
}

```

```{r include=FALSE}

# Function to plot confusion matrices 

plot_confusion_matrix <- function(conf_matrix, title) {
  conf_df <- as.data.frame(as.table(conf_matrix))
  colnames(conf_df) <- c("Predicted", "Actual", "Freq")
  
  ggplot(data = conf_df, aes(x = Predicted, y = Actual, fill = Freq)) +
    geom_tile(color = "black") +
    geom_text(aes(label = Freq), vjust = 1) +
    scale_fill_gradient(low = "white", high = "red") +
    labs(title = title, x = "Predicted", y = "Actual") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5),
          legend.position = "none")
}

```

In this section, we explore various models and techniques in order to train the best possible model with the goal to predict consumer churn. Since in general the oversampled dataset provided better performances, we are going to consider only the latter, except for cases in which certain results are particularly relevant.

#### Linear Models {.tabset .tabset-fade} 

In this section, we focus on building linear models to predict customer churn. In spite of the fact that the performance is highly affected by the linearity assumption, those remain fundamental since they serves as a benchmark for more complex models and helps validate the data and approach.
Subsequently, we perform variable selection, a crucial step in building efficient and interpretable models. It involves identifying the most relevant predictors from a set of potential variables. Two commonly used criteria for model selection are the Akaike Information Criterion (AIC) and the Bayesian Information Criterion (BIC).

##### Logistic Regression

In the first place, we consider as baseline the models in which no variable is used and all variables are included. Firstly, we build the model with all the variables.

```{r warning=FALSE}
set.seed(1)
model_baseline_logistic_oversample = glm(Churn ~ ., 
                              data = train_data_scaled_oversample, 
                              family = "binomial")

```

```{r include=FALSE}

# Storing the results in a confusion matrix
probabilities_baseline_logistic_oversample = predict(model_baseline_logistic_oversample, 
                                                          newdata = test_data_scaled_oversample,
                                                   type = "response")
predictions_baseline_logistic_oversample = ifelse(probabilities_baseline_logistic_oversample > 0.5, "True", "False")
confusion_matrix_baseline_logistic_oversample = table(predictions_baseline_logistic_oversample, 
                                                      test_data_scaled_oversample$Churn)

```

Then, we consider the model without any variable:

```{r}
model_0 = glm(Churn ~ 1,
                   family = "binomial",
                   data = train_data_scaled_oversample)

```

And subsequently we test the hypothesis of equivalence between the two models

```{r}

anova(model_0, model_baseline_logistic_oversample, test = "Chisq")

```

The null model fits the data poorly with a deviance of 5905.6. The full model, which includes all the specified predictors, fits the data significantly better, with a deviance of 4437.3. The reduction in deviance (1468.3) is highly significant, with a p-value \< 2.2e-16. Including the predictors in Model 2 significantly improves the fit of the model, indicating that the predictors collectively provide valuable information in predicting customer churn.


##### AIC

The AIC metric is used to compare models built with different set of predictors, with a focus on finding the model that best balances fit and complexity. It penalizes the number of parameters, discouraging overfitting, while prioritizing models with a better fit to the data. We performed feature selection using three different methods: forward selection, backward selection, and both directions selection.

```{r include=FALSE}
set.seed(1)
aic_model_forward_oversample = step(glm(Churn ~ 1, 
                      family = "binomial", 
                      data = test_data_scaled_oversample), 
                  scope = formula(model_baseline_logistic_oversample), 
                  direction = "forward")
aic_model_backward_oversample = step(model_baseline_logistic_oversample, direction = "backward")
aic_model_both_oversample = step(model_baseline_logistic_oversample, direction = "both")

# Number of features selected by each model
num_features_aic_forward_oversample = length(coef(aic_model_forward_oversample)) - 1
num_features_aic_backward_oversample = length(coef(aic_model_backward_oversample)) - 1
num_features_aic_both_oversample = length(coef(aic_model_both_oversample)) - 1

# Selected variables
selected_variables_aic_forward_oversample = names(coef(aic_model_forward_oversample))[-1]
selected_variables_aic_backward_oversample = names(coef(aic_model_backward_oversample))[-1]
selected_variables_aic_both_oversample = names(coef(aic_model_both_oversample))[-1]

```

```{r echo=FALSE}

# Print the number of features selected and the selected variables for each method
cat("With forward selection", num_features_aic_forward_oversample, "features have been selected.")
cat("\n")
print(paste(selected_variables_aic_forward_oversample, collapse = ", "))
cat("\n")

cat("With backward selection", num_features_aic_backward_oversample, "features have been selected")
cat("\n")
print(paste(selected_variables_aic_backward_oversample, collapse = ", ")) 
cat("\n")

cat("With both directions selection", num_features_aic_both_oversample, "features have been selected")
cat("\n")
print(paste(selected_variables_aic_both_oversample, collapse = ", ")) 
cat("\n")


```

Given the goal of creating a simpler model to avoid overfitting and enhance interpretability, we will choose the model generated by forward selection. This model includes only 9 features, providing a balance between simplicity and predictive power.

```{r include=FALSE}
# Select the simplest model, hence the one with the lowest features selected
feature_counts_oversample = c(forward = num_features_aic_forward_oversample, 
                              backward = num_features_aic_backward_oversample, 
                              both = num_features_aic_both_oversample)
selected_model_name_oversample = names(which.min(feature_counts_oversample))

if (selected_model_name_oversample == "forward") {
  model_aic_final_oversample = aic_model_forward_oversample
} else if (selected_model_name_oversample == "backward") {
  model_aic_final_oversample = aic_model_backward_oversample
} else {
  model_aic_final_oversample = aic_model_both_oversample
}

# Computing and storing predictions on the validation data
# Computing and storing predictions on the test data
probabilities_aic_oversample = predict(model_aic_final_oversample, test_data_scaled_oversample, type = "response")
predictions_aic_oversample = ifelse(probabilities_aic_oversample> 0.5, "True", "False")
confusion_matrix_aic_oversample = table(predictions_aic_oversample, test_data_scaled_oversample$Churn)

```

##### BIC

BIC is similar to AIC but applies a stronger penalty for the number of parameters. It is particularly useful for selecting simpler models, as it penalizes model complexity more heavily than AIC.

```{r include=FALSE}
set.seed(1)
bic_model_forward_oversample = step(glm(Churn ~ 1, 
                                        family = "binomial", 
                                        data = test_data_scaled_oversample), 
                                    scope = formula(model_baseline_logistic_oversample), 
                                    direction = "forward",
                                    k = log(nrow(test_data_scaled_oversample)))
bic_model_backward_oversample = step(model_baseline_logistic_oversample, direction = "backward", k = log(nrow(test_data_scaled_oversample)))
bic_model_both_oversample = step(model_baseline_logistic_oversample, direction = "both", k = log(nrow(test_data_scaled_oversample)))

# Number of features selected by each model
num_features_bic_forward_oversample = length(coef(bic_model_forward_oversample)) - 1
num_features_bic_backward_oversample = length(coef(bic_model_backward_oversample)) - 1
num_features_bic_both_oversample = length(coef(bic_model_both_oversample)) - 1

# Selected variables
selected_variables_bic_forward_oversample = names(coef(bic_model_forward_oversample))[-1]
selected_variables_bic_backward_oversample = names(coef(bic_model_backward_oversample))[-1]
selected_variables_bic_both_oversample = names(coef(bic_model_both_oversample))[-1]

```

```{r echo=FALSE}

# Print the number of features selected and the selected variables for each method
cat("With forward selection (BIC)", num_features_bic_forward_oversample, "features have been selected.")
cat("\n")
print(paste(selected_variables_bic_forward_oversample, collapse = ", "))
cat("\n")

cat("With backward selection (BIC)", num_features_bic_backward_oversample, "features have been selected")
cat("\n")
print(paste(selected_variables_bic_backward_oversample, collapse = ", "))
cat("\n")

cat("With both directions selection (BIC)", num_features_bic_both_oversample, "features have been selected")
cat("\n")
print(paste(selected_variables_bic_both_oversample, collapse = ", "))
cat("\n")

```

Again, the simplest model (the one with the lowest number of covariates) should be selected. In this case, with forward selection BIC selected 7 features confirming that it applies a stronger penalty for model complexity, often leading to simpler models.

In general, the process of feature selection confirmed that the variables International Plan, Total day minutes and Customer Service calls are relevant in order to predict customer Churn. Also, considering the stricter model we discover that the Region (State) categorical variable is not taken into account, while Voice mail plan, Total eve minutes, Total night minutes and Total intl minutes are kept in the model. The same features are selected by AIC that also considers Total intl calls and Total day calls.

```{r include=FALSE}

# Select the simplest model, hence the one with the lowest features selected
feature_counts_bic_oversample = c(forward = num_features_bic_forward_oversample, 
                                  backward = num_features_bic_backward_oversample, 
                                  both = num_features_bic_both_oversample)
selected_model_name_bic_oversample = names(which.min(feature_counts_bic_oversample))

if (selected_model_name_bic_oversample == "forward") {
  model_bic_final_oversample = bic_model_forward_oversample
} else if (selected_model_name_bic_oversample == "backward") {
  model_bic_final_oversample = bic_model_backward_oversample
} else {
  model_bic_final_oversample = bic_model_both_oversample
}

# Computing and storing predictions on the validation data
# Computing and storing predictions on the validation data
probabilities_bic_oversample = predict(model_bic_final_oversample, test_data_scaled_oversample, type = "response")
predictions_bic_oversample = ifelse(probabilities_bic_oversample > 0.5, "True", "False")
confusion_matrix_bic_oversample = table(predictions_bic_oversample, test_data_scaled_oversample$Churn)

```

##### Comparisson between Linear models

First, we report the confusion matrix for each of the three previously trained models to provide a general understanding of their performance. It is important to note that each prediction was made on the test set, which had been scaled using the parameters derived from the training set.

```{r echo=FALSE}

# Create the plots
baseline_plot = plot_confusion_matrix(confusion_matrix_baseline_logistic_oversample, "Baseline Model")
aic_plot = plot_confusion_matrix(confusion_matrix_aic_oversample, "AIC Model")
bic_plot = plot_confusion_matrix(confusion_matrix_bic_oversample, "BIC Model")

# Arrange the plots in a single layout
grid.arrange(baseline_plot, aic_plot, bic_plot, ncol = 3, top = textGrob("Confusion Matrices", gp = gpar(fontsize = 20, fontface = "bold")))

```

The Baseline Model has a surprisingly high number of true positives relative to the total actual positives, indicating it is particularly effective at minimizing false negatives. However, this effectiveness comes at the cost of a high number of false positives, as the model tends to classify many observations as positive. While this aligns with our aim of minimizing false negatives, the overall accuracy of the model is too low, indicating significant room for improvement. In contrast, the BIC and AIC models reduce misclassification errors but perform poorly in identifying true positives. This may be due to the models' excessive simplicity, which fails to capture certain patterns in the test data. The similarity in the features selected by the BIC and AIC models is evidenced by their almost identical correlation matrices.

Also, we consider their performance in terms of AUC (Area Under ROC Curve). The ROC (Receiver Operating Characteristic) curve is a graphical representation of a classifier's performance, plotting the true positive rate against the false positive rate at various threshold settings.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Compute ROC curves

library(pROC)
par(mfrow=c(1,3)) # 3 rows, 1 column layout for confusion matrices
roc_baseline = roc(test_data$Churn, 
                   predict(model_baseline_logistic_oversample, 
                           newdata = test_data_scaled_oversample, 
                           type = "response"), 
                   plot = TRUE, main = "Baseline Model", col = "blue", lwd = 3, 
                   auc.polygon = TRUE, print.auc = TRUE)
roc_aic = roc(test_data$Churn, 
                   predict(model_aic_final_oversample, 
                           newdata = test_data_scaled_oversample, 
                           type = "response"), 
                   plot = TRUE, main = "AIC Model", col = "red", lwd = 3, 
                   auc.polygon = TRUE, print.auc = TRUE)
roc_bic = roc(test_data$Churn, 
              predict(model_bic_final_oversample, 
                      newdata = test_data_scaled_oversample, 
                      type = "response"), 
              plot = TRUE, main = "BIC Model", col = "purple", lwd = 3, 
              auc.polygon = TRUE, print.auc = TRUE)
```


The Baseline Model shows decent performance with an AUC of 0.827, but it is outperformed by both the AIC and BIC models. Also, we can see that the specificity is particulary high even for high levels of sensitivity, confirming that the model is great at identifying true positives.
AIC and BIC models, as saw before, present a more balanced result. The AIC Model demonstrates the best performance with the highest AUC of 0.854, indicating a slightly higher discrimination capability between churn and non-churn customers. The BIC Model also performs well, with an AUC of 0.846, closely following the AIC Model and significantly better than the Baseline Model.


#### Penalized approaches {.tabset .tabset-fade} 

Penalized approaches like Lasso (Least Absolute Shrinkage and Selection Operator) and Ridge regression add regularization terms to the loss function to prevent overfitting. We decided to test these methods since AIC and BIC models showed minimal improvement with respect to the baseline model and hence a regularization approach could have enhanced model performance by managing complexity and selecting relevant features more effectively.

##### Lasso Regression

Lasso regression adds a penalty equal to the absolute value of the coefficients, encouraging sparsity by shrinking some coefficients to zero, effectively selecting only the most important predictors. This helps in building simpler and more interpretable models.

We start by training the model.
```{r echo=FALSE, warning=FALSE}
set.seed(1)
# Train the Lasso regression model on the oversampled data
X_train_oversample = train_data_scaled_oversample
Y_train_oversample = as.numeric(X_train_oversample$Churn)
X_train_oversample$Churn = NULL
fit_oversample = glmnet(as.matrix(X_train_oversample), Y_train_oversample, alpha = 1, lambda = seq(0, 0.15, length = 30))

```

Afterwards, we consider the relation between the coefficients and the parameter lambda.
```{r echo=FALSE, warning=FALSE}
set.seed(1)
# Plot coefficient values against the (log-)lambda sequence
predictor_names_oversample = colnames(as.matrix(X_train_oversample))
colors = c("black", "blue", "red", "yellow", "green", "orange", "purple", "deepskyblue",
           "grey", "#3D0C02", "violet", "#9EFD38", "#3B7A57", 	"#841B2D")
plot(fit_oversample, xvar = "lambda", label = TRUE, col = colors)
legend("topright", legend = predictor_names_oversample, col = colors, lty = 1, cex = 0.5, text.width = 1.2)

```

This plot shows how the coefficients of different features in a Lasso regression model change as the regularization parameter increases (left on the x-axis). As it increases, the coefficients shrink toward zero, demonstrating Lasso's feature selection capability. Features with coefficients that remain large even at higher values are considered to be more important, while those that quickly shrink to zero are less important. An interesting result is that the predictor that is shrinked only at higher values is "total international minutes".


After the preliminary analysis, we perform the crucial step of hyperparameter tuning to identify, with cross validation, an estimate for the best value for lamda.

```{r}
set.seed(1)
# Train the Lasso regression model on the oversampled data with tuning
model_lasso_oversample = train(Churn ~ ., 
                               data = train_data_scaled_oversample, 
                               method = "glmnet", 
                               metric = "Accuracy", 
                               trControl = trainControl(method = "cv", number = 10), 
                               tuneGrid = expand.grid(alpha = 1, lambda = seq(0, 0.15, length = 30)))


```

```{r echo=FALSE}
# Plot accuracy values against the lambda sequence
plot(model_lasso_oversample,
     label = TRUE, 
     xvar = "lambda")

# Retrieve the maximum accuracy and best tuning parameters 
# achieved during cross-validation
best_accuracy_lasso_oversample = max(model_lasso_oversample$results$Accuracy)
best_parameters_lasso_oversample = model_lasso_oversample$bestTune$lambda
cat("The highest value of accuracy:", best_accuracy_lasso_oversample, "is obtained with lambda =", best_parameters_lasso_oversample, "\n")

```

The best accuracy is obtained without a regularization parameter: in follows that the model built is a simple linear regression model. As the parameter continues to increase beyond 0.05, the accuracy starts to decline more steeply, indicating that the model is becoming overly regularized. This excessive penalization of the coefficients leads to underfitting, where the model is too simple to capture the underlying patterns in the data. The conclusion is similar with the one obtained with AIC and BIC: in order to have a satisfying prediction performance then each feature must be included in the model.

```{r include=FALSE}

# Computing and storing predictions
probabilities_lasso_oversample = predict(model_lasso_oversample, test_data_scaled_oversample, type = "prob")[, "True"]
predictions_lasso_oversample = ifelse(probabilities_lasso_oversample > 0.5, "True", "False")
confusion_matrix_lasso_oversample = table(predictions_lasso_oversample, test_data_scaled_oversample$Churn)

```

##### Ridge Regression

Ridge regression is another linear regression technique that includes a penalty term to reduce model complexity and prevent overfitting. Unlike Lasso regression, which can shrink some coefficients to zero and perform feature selection, with Ridge the coefficients tend to zero but they are never exactly to zero.

As we have done before, we firstly train the model to have an overview of the relation between the parameter lambda and the coefficients.

```{r}

set.seed(1)
ridge_fit_oversample = glmnet(x = X_train_oversample,
                              y = Y_train_oversample,
                              alpha = 0)
```

```{r echo=FALSE}
plot(ridge_fit_oversample,
     label = F, 
     xvar = "lambda", col = colors)
legend("topright", legend = predictor_names_oversample, col = colors, lty = 1, cex = 0.5, text.width = 1.2)

```

The plot confirms the earlier statement: as the parameter Lambda increases, the coefficients shrink towards zero but never actually reach it, in contrast with what we have previously shown with Lasso. 

We then proceed, as we have done for Lasso regression, to tune the parameters selecting the value with which the model reaches the highest accuracy on the validation set.

```{r}

set.seed(1)
model_ridge_oversample = train(Churn ~ ., 
                               data = train_data_scaled_oversample, 
                               method = "glmnet", 
                               metric = "Accuracy", 
                               trControl = trainControl(method = "cv", number = 10), 
                               tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 0.15, length = 30)))


```

```{r echo=FALSE}


model_ridge_oversample %>% 
  ggplot(aes(x = lambda, y = Accuracy)) + 
  geom_line() + 
  geom_point() +
  geom_text(aes(label = sprintf("%.3f", Accuracy)), check_overlap = TRUE, vjust = -0.5, size = 2.5) + 
  labs(x = TeX("Lambda ($\\lambda$)"), y = "Accuracy", title = "Accuracy vs. Lambda for Ridge Regularization") +
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )


cat("The highest accuracy is:", max(model_ridge_oversample$results$Accuracy),
    "with parameter lambda =", model_ridge_oversample$bestTune$lambda)


```

```{r include=FALSE}

probabilities_ridge_oversample = predict(model_ridge_oversample, test_data_scaled_oversample, 
                                         type = "prob")[,"True"]
predictions_ridge_oversample = ifelse(probabilities_ridge_oversample > 0.5, "True", "False")
confusion_matrix_ridge_oversample = table(predictions_ridge_oversample, test_data_scaled_oversample$Churn)

```

##### Comparisson between penalized approaches

```{r echo=FALSE}

## Confusion matrices

# Create the plots
lasso_plot = plot_confusion_matrix(confusion_matrix_lasso_oversample, "Lasso")
ridge_plot = plot_confusion_matrix(confusion_matrix_ridge_oversample, "Ridge")


# Arrange the plots in a single layout
grid.arrange(lasso_plot, ridge_plot, ncol = 2, top = textGrob("Confusion Matrices", gp = gpar(fontsize = 20, fontface = "bold")))


```

The similarity between the confusion matrices for the penalized approaches (Lasso and Ridge) and the baseline model indicates that the lambda parameter has little impact on the results. In the case of Lasso, the lambda value is almost negligible, and for Ridge, it is particularly small. This confirms that removing and shrinking features does not benefit our analysis. Consequently, these three methods essentially represent the same model, leading us to decide not to report the AUC curves for the penalized approaches. Including them in our final conclusions would be redundant, as the findings would merely reiterate what has already been discussed regarding the baseline model's performance.

```{r echo=FALSE, warning=FALSE, message=FALSE, include = FALSE}

## ROC

# Compute ROC curves

library(pROC)
par(mfrow=c(1,2)) # 3 rows, 1 column layout for confusion matrices
roc_lasso = roc(test_data$Churn, probabilities_lasso_oversample, 
             plot = TRUE, main = "Lasso", col = "blue", lwd = 3, 
             auc.polygon = TRUE, print.auc = TRUE)
roc_aic = roc(test_data$Churn, 
              probabilities_ridge_oversample, 
              plot = TRUE, main = "Ridge", col = "red", lwd = 3, 
              auc.polygon = TRUE, print.auc = TRUE)
par(mfrow=c(1,1))


```

#### Non linear methods {.tabset .tabset-fade}

While linear and penalized regression methods provide a solid foundation for understanding the relationships between predictors and the response variable, they often fall short when dealing with more complex, non-linear interactions within the data. To address this, we turn to non-linear methods which can capture intricate patterns and improve predictive performance. Non-linear models, such as the decision tree seen before, random forests, and boosting techniques, offer flexibility in modeling relationships that linear methods cannot adequately describe.

##### Random Forests

With Random Forests, a number of decision trees is built on bootstrapped training samples. However, each time a split in a tree is considered, a random sample of m predictors is chosen as split candidates from the full set of predictors.

We first start by building a simple model with default parameters. The number of predictors considered at each split (mtry in the code) is typically the square root of the number of the predictors. In our case is then 4. We also consider a forest with 500 trees.

```{r}

set.seed(1)
random_forest_model_oversample = randomForest(Churn ~ ., data = train_data_oversample, 
                                              mtry = 4, 
                                              ntree = 500, 
                                              importance = T)
```

In this initial analysis, our primary focus is not on model performance but on observing how the Out-Of-Bag (OOB) error changes as we increase the number of trees.

```{r echo=FALSE}

# Storing the error rate matrix
error_rate_matrix_oversample = random_forest_model_oversample$err.rate

# Creating the error rate data frame for plotting
error_rate_matrix_oversample = data.frame(
  Trees = rep(1:nrow(error_rate_matrix_oversample), times = 3),
  Type = rep(c("OOB", "False", "True"), each = nrow(error_rate_matrix_oversample)),
  Error = c(error_rate_matrix_oversample[, "OOB"],
            error_rate_matrix_oversample[, "False"],
            error_rate_matrix_oversample[, "True"])
)

# Plotting the error rate
ggplot(data = error_rate_matrix_oversample, aes(x = Trees, y = Error)) +
  geom_line(aes(color = Type))

```

We can see that the out-of-bag (OOB) error stabilizes after around 100 trees, indicating that the random forest model achieves consistent performance. The error for the "True" class is significantly lower than the "False" class, suggesting the model is better at predicting "True" than "False". This could be caused by the fact that we are training the model on an oversampled balanced data. However, we also tested the model on unbalanced data and its performance was significantly worse.

We can also show the features importance considering the metric of mean decrease accuracy (computed by measuring the increase in the model's OOB error when a feature's values are randomly permuted; higher decreases in accuracy indicate greater importance).

```{r echo=FALSE}

importance_values = importance(random_forest_model_oversample)

# Plotting the feature importance plot for MeanDecreaseAccuracy 
importance_df = data.frame(
  Feature = rownames(importance_values),
  MeanDecreaseAccuracy = importance_values[, "MeanDecreaseAccuracy"]
)

importance_df = importance_df[order(importance_df$MeanDecreaseAccuracy, decreasing = TRUE), ]
par(mar = c(5, 8, 4, 2) + 0.1)  
barplot(
  importance_df$MeanDecreaseAccuracy,
  names.arg = importance_df$Feature,
  las = 2,
  col = "skyblue",
  main = "Feature Importance - Mean Decrease in Accuracy",
  xlab = "Mean Decrease in Accuracy",
  cex.names = 0.7, 
  horiz = TRUE
)
par(mar = c(5, 4, 4, 2) + 0.1) 

```

We can see that "Total day minutes", "Customer service calls", "Total eve minutes" and "International Plan" are among the most important features, as they have the highest values in both metrics, indicating their strong influence on model accuracy and confirming our findings so far.


One of the most straightforward drawbacks of random forests is their tendency to overfit the data due to the inherent nature of decision trees, which are particularly prone to overfitting. Although random forests mitigate this issue by increasing model variability, hyperparameter tuning is essential for optimal performance.

In our R reference code, we employed two different strategies to tune the parameters. First, we used a classic approach with the "train" function, specifying a tune grid to test with 5-fold cross-validation. However, this function automatically tunes only the parameter mtry and does not consider other relevant parameters such as the number of trees, maximum number of nodes, and node size. Therefore, we built a grid, randomly sampled combinations, and tested them using the OOB error, which is an accurate estimate of the test error. Since this approach yielded better results, we decided to report it.

```{r include=FALSE}

# Create a random sample of hyperparameter combinations
set.seed(1)
hyperparameter_grid <- expand.grid(
  n_tree = c(400,450,500,550,600),
  m_try = 1:12,
  max_nodes = c(10, 25, 50, 75, 100, 200, 300),
  node_size = 1:10
)

# Sample a smaller grid for random search
sampled_grid = hyperparameter_grid %>% sample_n(50)

# Evaluate all combinations
results_random_forest_tuning = data.frame()
for (i in 1:nrow(sampled_grid)) {
  params = sampled_grid[i, ]
  print(params)
  rf_model = randomForest(
    Churn ~ .,
    data = train_data_oversample,
    ntree = params$n_tree,
    mtry = params$m_try,
    maxnodes = params$max_nodes,
    nodesize = params$node_size,
    importance = TRUE
  )
  # Calculate the mean error rate
  mean_err_rate = mean(rf_model$err.rate[, 1])
  print(mean_err_rate)
  
  # Appending the results in the dataframe
  results_random_forest_tuning = rbind(results_random_forest_tuning,
                                       cbind(params, mean_err_rate))
}

```

```{r echo=FALSE}

# Find the best parameters
best_parameters_random_forest = results_random_forest_tuning %>%
  filter(mean_err_rate == min(mean_err_rate))

kable(best_parameters_random_forest, caption = "Best Parameters for Random Forest")

```

Once we have found our tuned hyperparameters, we use them in order to build the model that will be used in order for us to predict Customer Churn.

```{r}

# Train the final model with the best parameters
final_rf_model <- randomForest(
  Churn ~ .,
  data = train_data_oversample,
  ntree = best_parameters_random_forest$n_tree,
  mtry = best_parameters_random_forest$m_try,
  nodesize = best_parameters_random_forest$node_size,
  maxnodes = best_parameters_random_forest$max_nodes,
  importance = TRUE
)

```

```{r include=FALSE}

# Make predictions on the test set
probabilities_rf = predict(final_rf_model, test_data, type = "prob")[, "True"]
predictions_rf = ifelse(probabilities_rf > 0.5, "True", "False")
confusion_matrix_rf = table(predictions_rf, test_data$Churn)

```

##### XGBOOST

XGBoosting is an advanced implementation of gradient boosting that builds an ensemble of decision trees sequentially, with each tree correcting the errors of the previous ones, optimized for speed and performance.

Since XGBoost requires numerical input and cannot directly handle categorical variables, as a first essential step we created dummy variables to encode categorical variables.

```{r warning=FALSE, include=FALSE}

# Create dummy variables for the data
X_train_oversample_dummy = dummyVars(Churn ~ .
                                     , data = train_data_oversample) %>% predict(train_data_oversample)
Y_train_oversample_dummy = train_data_oversample$Churn %>% as.numeric(.) - 1
X_test_oversample_dummy = dummyVars(Churn ~ ., data = test_data) %>% predict(test_data)
Y_test_oversample_dummy = test_data$Churn %>% as.numeric(.) - 1

```

Then, in order to have a general understanding of the performance of the model, we trained it without tuning the hyperparameters.

```{r message=FALSE, results='hide'}

# In the first place we train the algorithm without tuning
set.seed(1)
model_xg = xgboost(as.matrix(X_train_oversample_dummy), 
                   label = Y_train_oversample_dummy, 
                 nrounds = 50, 
                 objective = "binary:logistic", 
                 eval_metric = "error")


```

To gain insight into the relevance of the ntreelimit parameter (and thereby the importance of tuning hyperparameters), we consider the error on the test set. It is important to note that this analysis is not part of the model training process but is merely an evaluation of the parameter, as models should not be trained on the test set.

```{r echo=FALSE}

train_errors = model_xg$evaluation_log$train_error
val_errors = numeric(50)

for (j in 1:50) {
  pred_j = ifelse(predict(model_xg, X_test_oversample_dummy)> 0.5, 1, 0)
  val_errors[j] <- mean(pred_j != Y_test_oversample_dummy)
}

# Plot the error rates
plot(1:50, val_errors, type = "b", xlab = "Number of trees", ylab = "Error", col = 3, ylim = c(0, 0.3), cex = 0.5)
points(1:50, train_errors, type = "b", cex = 0.5)
legend("topright", legend = c("Train", "Test"), col = c(1, 3), lty = 1, lwd = 2, cex = 0.7)

```

```{r include=FALSE}

# We then consider its performance
probabilities_xg = predict(model_xg, X_test_oversample_dummy)
predictions_xg = ifelse(probabilities_xg> 0.5, 1, 0)
confusion_matrix_xg = table(predictions_xg, test_data$Churn)

```

The plot shows that as the number of trees increases, the training error consistently decreases, approaching zero, indicating that the model fits the training data very well. However, the test error stabilizes at a low value early on and remains relatively constant, suggesting that additional trees do not significantly improve the model's performance on the test data. This indicates that the model generalizes well without overfitting, as there is a minimal gap between the training and test errors after a certain number of trees.

We then proceed with the process of hyperparameter tuning using cross validation. Since tasks like cross-validation can be paralized, in order to speed up computations, we created a cluster of 5 worker nodes with the library doParallel.

```{r}

cl = makePSOCKcluster(5)
registerDoParallel(cl)

```

```{r include=FALSE}

tune_grid = expand.grid(
  nrounds = c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100),
  eta = 0.3,
  max_depth = c(1,2,3,4,5,6,7,8),
  subsample = 1,
  colsample_bytree = 1,
  min_child_weight = 5,
  gamma = 0 #c(0.1, 0.2, 0.5, 0.75, 1)
)

fitControl = trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 10,
  search = "random"
)

```

```{r warning=FALSE}

set.seed(1)
model_xg_cv = train(
  Churn ~ ., data = train_data_oversample, 
  method = "xgbTree", 
  trControl = fitControl,
  verbose = FALSE, 
  tuneGrid = tune_grid,
  objective = "binary:logistic", 
  eval_metric = "error"
)

```

```{r echo=FALSE, warning=FALSE}

# Stop parallel processing
stopCluster(cl)

# Plot the cross-validation results
trellis.par.set(caretTheme())
plot(model_xg_cv)

# Print best parameters
best_params_df = as.data.frame(model_xg_cv$bestTune)
kable(best_params_df, caption = "Best Parameters for XGBOOST")

```

The plot shows that increasing the maximum tree depth significantly improves model accuracy, with depths of 7 and 8 performing the best, achieving over 95% accuracy. The accuracy improves with the number of boosting iterations, plateauing around 60-80 iterations for depths 5 and above, indicating diminishing returns beyond this point. Lower depths (1-3) exhibit significantly lower accuracy, staying below 85%. To optimize performance, it is crucial to tune both the number of boosting iterations and the maximum tree depth, favoring higher depths (but we need to take into account the risk of overfitting).

```{r include=FALSE}

# Predictions on the test set using the best model
probabilities_xg_cv = predict(model_xg_cv, test_data, type = "prob")[,"True"]
predictions_xg_cv = ifelse(probabilities_xg_cv > 0.5, "True", "False")
confusion_matrix_xg_cv = table(predictions_xg_cv, test_data$Churn)

```

##### Comparisson between Tree based models

As we have done before, to firstly asses the performances we plot the confusion matrices for Random Forests and XGBOOST (with and without tuning).

```{r echo=FALSE}

## Confusion matrices

# Create the plots
rf_plot = plot_confusion_matrix(confusion_matrix_rf, "Random Forest")
xg_plot = plot_confusion_matrix(confusion_matrix_xg, "XGBOOST")
xg_cv_plot = plot_confusion_matrix(confusion_matrix_xg_cv, "XGBOOST (Tuned)")

# Arrange the plots in a single layout
grid.arrange(rf_plot, xg_plot, xg_cv_plot, ncol = 3, top = textGrob("Confusion Matrices", gp = gpar(fontsize = 20, fontface = "bold")))

```

The initial belief that non-linear models would perform better is confirmed. The tree-based methods present similar confusion matrices, balancing sensitivity and specificity: indeed, the number of true positives is significantly higher than in the previous cases without any tradeoff in overall accuracy. Furthermore, the models excel at classifying the negative class, with very few misclassifications.

A peculiar result is that the untuned model seems to perform better than the tuned model. However, since the difference between the two is only three observations, this is likely due to the way the test and train sets were split. It would be inaccurate to conclude definitively that the untuned model is better, but also that this result is purely by chance. Indeed, it is possible that better hyperparameter tuning (requiring higher computational power) is needed or that the cross-validation error did not provide an accurate estimate of the test error.

The substantial increase in performance is confirmed also by the AUC metrics:

```{r echo=FALSE, warning=FALSE, message=FALSE}

## ROC

# Compute ROC curves

library(pROC)
par(mfrow=c(1,3)) # 3 rows, 1 column layout for confusion matrices
roc_rf = roc(test_data$Churn, probabilities_rf, 
                   plot = TRUE, main = "Random Forest", col = "blue", lwd = 3, 
                   auc.polygon = TRUE, print.auc = TRUE)
roc_aic = roc(test_data$Churn, 
              probabilities_xg, 
              plot = TRUE, main = "XGBOOST", col = "red", lwd = 3, 
              auc.polygon = TRUE, print.auc = TRUE)
roc_bic = roc(test_data$Churn, 
              probabilities_xg_cv, 
              plot = TRUE, main = "XGBOOST (Tuned)", col = "purple", lwd = 3, 
              auc.polygon = TRUE, print.auc = TRUE)
par(mfrow=c(1,1))


```

### CONCLUSIONS

Given the high imbalance in our dataset, we have prioritized sensitivity in our selection criteria because minimizing false negatives and thereby reducing missed positives is crucial for our analysis. In the telecom industry, failing to predict a customer’s churn (false negative) is more costly than predicting that a customer will churn when they won’t (false positive). This approach helps ensure we capture as many potential churners as possible, enabling proactive retention efforts and ultimately reducing customer attrition. However, balancing sensitivity with accuracy remains important: if there are too many false positives because the models tend to classify most observations as positive, proactive retention efforts may become too costly and often unnecessary. Therefore, while sensitivity (recall) is our primary focus to mitigate the higher cost of missed churn predictions, the overall accuracy must also be satisfactory to ensure a feasible and effective retention strategy.

```{r echo=FALSE}
# Calculate metrics for each model
metrics_baseline = get.metrics(confusion_matrix_baseline_logistic_oversample)
metrics_aic = get.metrics(confusion_matrix_aic_oversample)
metrics_bic = get.metrics(confusion_matrix_bic_oversample)
metrics_rf = get.metrics(confusion_matrix_rf)
metrics_xg = get.metrics(confusion_matrix_xg)
metrics_xg_cv = get.metrics(confusion_matrix_xg_cv)

# Combine metrics into a single data frame
metrics_combined = rbind(metrics_baseline, metrics_aic, metrics_bic,
                         metrics_rf, metrics_xg, metrics_xg_cv)
rownames(metrics_combined) = c("Baseline", "AIC", "BIC",
                               "Random Forests", "XGBOOST", "XGBOOST Tuned")

# Round the metrics to 2 decimal places
metrics_combined = round(metrics_combined, 4)

kable(metrics_combined)

```
Considering our initial goal of minimizing false negatives, the best model appears to be the XGBOOST with default parameters. However, given the minimal difference in performance, if we were to predict customer churn on unobserved data, we believe the tuned XGBOOST model has the highest probability of delivering the best performance. This is because the results on the test set can be heavily influenced by the specific nature of the observations, and changing the random seed could lead to different outcomes. Nonetheless, hyperparameter tuning remains a crucial technique that must be applied to tailor the model to the specific problem we are addressing.

Overall, we are satisfied with the performance of the non-linear methods, as evidenced by their balanced sensitivity and specificity. The high overall accuracy is further supported by elevated F1 scores, which require both high recall (sensitivity) and precision. These metrics demonstrate that our models are not only accurate but also effective at identifying true positives and minimizing false positives, making them robust choices for predicting customer churn.


## TASK 2 {.tabset .tabset-fade}

For this task, we provided a customer segmentation which is based on the CLV index (Customer Lifetime Value) and provides insights about groups of customers that the company can target with ad-hoc promotional campaigns. In this way the company would be more efficient in targeting only customers who have a large CLV.

```{r include=FALSE}
# Importing again the dataset since it was cleaned

Data = read.csv2("./Dataset/TelecomChurn.csv", 
                 header = T, 
                 sep = ",", 
                 colClasses = "character")
variables = colnames(Data)
categorical_variables = c("State", "International.plan", "Voice.mail.plan", "Area.code")
target_variable = "Churn"
numerical_variables = setdiff(variables, c(categorical_variables, target_variable))
predictors = setdiff(variables, target_variable)

# Ensuring that variables are converted in the correct form
Data[[target_variable]] = as.factor(Data[[target_variable]])
for (var in numerical_variables) {
  Data[[var]] = as.numeric(Data[[var]])
}
for (var in categorical_variables) {
  Data[[var]] = as.factor(Data[[var]])
}



```


### CLV Computation and Distribution

Since our dataset did not include relevant features that could have been used for an insightful interpretation (since the dataset is mainly provided in order to train models to predict customer churn) we decided to aproximate the CLV. Customer Lifetime Value (CLV) is vital for predicting revenue, allocating resources efficiently, and segmenting customers for targeted marketing. It helps manage acquisition costs by ensuring they do not exceed the customer's lifetime value and supports strategic decision-making in product development, pricing, and customer relationship management. 

#### Computing CLV

In the first place, we computed "Total charge" in order to estimate the revenue for each customer. 
The average customer lifespan is another important metric needed to compute the CLV. We decided to compute it based on the assumption that if a fraction of customers churn each period, then, on average, a customer will stay for 1/churn rate periods. A lower Churn Rate indicates indeed an higher Customer Lifespan.
We then proceeded by computing (approximating) the CLV as the product between the Total Charges for each customer and the average customer lifetime span.
Ultimately, we calculated the total daily minutes as an additional "summary" metric representing the time a customer spends using the service each day. The creation of this index serves to enhance and expand our set of features for clustering analysis. By including total daily minutes, we aim to capture a more comprehensive picture of customer behavior, allowing for more accurate and insightful clustering results.

```{r }

Data$Total.charge = Data$Total.day.charge + Data$Total.eve.charge + 
  Data$Total.night.charge + Data$Total.intl.charge #Total Charge

churn_rate = mean(Data$Churn == "True") # Churn Rate
retention_rate = 1 - churn_rate #Customer Retention Rate. 
avg_customer_lifespan = 1 / churn_rate #AVG Customer Lifetime span

Data$CLV = Data$Total.charge * avg_customer_lifespan

Data$Total.minutes = Data$Total.day.minutes + Data$Total.eve.minutes + 
  Data$Total.night.minutes + Data$Total.intl.minutes #Total Minutes

```

####  CLV Distribution

Now that we have computed the CLV index, which is basically a summary of the various charges paid by each customer, we analyze it like we did for the other features in the EDA to gain insight about its distribution and its relation with some other features.

```{r echo=FALSE}
ggplot(Data, aes(x = CLV)) +
  geom_histogram(binwidth = 50, fill = "#0073C2FF", color = "black", alpha = 0.7) +
  theme_minimal(base_size = 15) +
  labs(
    title = "Distribution of Customer Lifetime Value (CLV)",
    x = "Customer Lifetime Value (CLV)",
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  )
```

The histogram shows that most customers have a Customer Lifetime Value (CLV) centered around 400 units, with the majority of values falling between 300 and 500. The distribution is slightly left-skewed, indicating some customers have significantly lower CLV. This central tendency suggests a typical customer profile, while the skew highlights opportunities for targeted retention strategies. Overall, the distribution provides insights for optimizing customer value and retention efforts.

##### CLV by Top Ten States

Considering the dataset, understand if the CLV were distributed differently along states is an essential analysis to perform.

```{r include=FALSE}
# Identify the top ten most common states
top_states <- Data %>%
  count(State, sort = TRUE) %>%
  top_n(10, wt = n) %>%
  pull(State)

# Filter the data to include only the top ten states
filtered_data <- Data %>% filter(State %in% top_states)

```

```{r echo=FALSE}
# CLV by State for Top Ten States
ggplot(filtered_data, aes(x = State, y = CLV)) +
  geom_boxplot(fill = "#0073C2FF", alpha = 0.7, outlier.color = "red", outlier.shape = 16) +
  theme_minimal(base_size = 15) +
  labs(
    title = "CLV by State (Top Ten States)",
    x = "State",
    y = "Customer Lifetime Value (CLV)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

In general, the median CLV values are relatively similar across the states, indicating a consistent customer value distribution. However, some states exhibit greater variability, as evidenced by wider IQRs and more outliers.


### CLUSTERING ALGORITHMS {.tabset .tabset-fade}

In this section, we develop clustering algorithms to analyze customer segments. We will explore two clustering techniques: K-means and Hierarchical clustering. For this analysis, we focus on the features Account Length, Total Minutes, and Customer Lifetime Value (CLV). These features are selected because they are highly representative and particularly relevant for our objective of understanding customer behavior and optimizing retention strategies. By leveraging these clustering techniques, we aim to identify distinct customer groups that can inform targeted marketing and retention efforts.
Also, a crucial step is to perform scaling on the dataset.


``` {r}

# Selection and normalization of the features
clustering_data = Data[, c("Account.length", "CLV", "Total.minutes")]
clustering_data_scaled = scale(clustering_data)

```

#### K-Means

Firstly, we tried to segment customers with K-means clustering, a popular and efficient method for partitioning a dataset into distinct groups or clusters. K-means clustering aims to minimize the variance within each cluster, thereby maximizing the homogeneity of data points within clusters and the heterogeneity between clusters.

##### Identification of optimal number of clusters

We determined the optimal number of clusters using both the elbow method and the silhouette method. The elbow method identified the point where the within-cluster sum of squares (WSS) sharply decreases, indicating the optimal balance between model complexity and accuracy. The silhouette method measured how well data points fit within their assigned clusters compared to others. By comparing both methods, we ensured a robust and well-defined clustering solution.

```{r include= FALSE, warning=FALSE}

set.seed(1)
wss = sapply(1:10, function(k) {
  kmeans(clustering_data_scaled, centers = k, nstart = 10)$tot.withinss
})

silhouette_width = sapply(2:10, function(k) {
  km = kmeans(clustering_data_scaled, centers = k, nstart = 10)
  ss = silhouette(km$cluster, dist(clustering_data_scaled))
  mean(ss[, 3])
})

```

```{r echo=FALSE, fig.width= 10}

# Storing results
k_means_elbow_method = data.frame(Clusters = 1:10, WSS = wss)
k_means_silhouette_method = data.frame(Clusters = 2:10, Silhouette = silhouette_width)

# Function to plot results
plot_elbow_method = function(df, optimal_k) {
  ggplot(df, aes(x = Clusters, y = WSS)) +
    geom_line(color = "#0073C2FF", size = 1.2) +
    geom_point(color = "#0073C2FF", size = 3) +
    labs(title = "Elbow Method",
         x = "Number of Clusters",
         y = "Total Within-Clusters Sum of Squares") +
    theme_minimal(base_size = 15) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold", color = "black"),
      axis.title = element_text(size = 16, face = "bold", color = "black"),
      axis.text = element_text(size = 14, face = "bold", color = "black"),
      panel.grid.major = element_line(color = "grey80"),
      panel.grid.minor = element_line(color = "grey90"),
      panel.background = element_rect(fill = "whitesmoke", color = NA)
    )
}

# Function to plot Silhouette method
plot_silhouette_method = function(df) {
  ggplot(df, aes(x = Clusters, y = Silhouette)) +
    geom_line(color = "#D9534F", size = 1.2) +
    geom_point(color = "#D9534F", size = 3) +
    labs(title = "Silhouette Method",
         x = "Number of Clusters",
         y = "Average Silhouette Width") +
    ylim(0, 0.4) +
    theme_minimal(base_size = 15) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold", color = "black"),
      axis.title = element_text(size = 16, face = "bold", color = "black"),
      axis.text = element_text(size = 14, face = "bold", color = "black"),
      panel.grid.major = element_line(color = "grey80"),
      panel.grid.minor = element_line(color = "grey90"),
      panel.background = element_rect(fill = "whitesmoke", color = NA)
    )
}

plot_elbow = plot_elbow_method(k_means_elbow_method)
plot_siluette = plot_silhouette_method(k_means_silhouette_method)

grid.arrange(plot_elbow, plot_siluette, ncol = 2)

```


The Elbow Method plot shows the within-cluster sum of squares (WSS) against the number of clusters. As the number of clusters increases, the WSS decreases, indicating improved cluster compactness. The elbow point, where the rate of decrease sharply slows, suggests 2 clusters as an optimal balance between model complexity and accuracy.
The Silhouette Method plot displays the average silhouette width against the number of clusters. The highest silhouette width is also observed at 2 clusters, indicating the best fit and most distinct cluster separation at this point.

Although both methods indicate that 2 clusters are optimal, we chose to select 3 clusters. At 3 clusters, the WSS value is further minimized, and we are not overly concerned about overfitting. Additionally, selecting 3 clusters provides a more flexible and explorative framework, allowing us to uncover hidden patterns and structures in the data. This approach helps us identify distinct customer segments and gain a deeper understanding of the data.


##### K-Means in action

Once we have found the optimal value for k, it is possible to apply the k-means algorithm in order to to actually find and visualize customer segments (with dimensionality reduction).

```{r}
# Apply K-means clustering with the chosen number of clusters 
set.seed(1)
kmeans_result = kmeans(clustering_data_scaled, centers = 3, nstart = 25)

```

```{r echo=FALSE}
# Append cluster results to the original data
Data$Cluster = as.factor(kmeans_result$cluster)

# Visualize the clusters
fviz_cluster(kmeans_result, data = clustering_data_scaled, geom = "point", ellipse.type = "convex", palette = "jco", ggtheme = theme_minimal())

```

The cluster plot shows the results of K-means clustering, displaying the distribution of data points across nine distinct clusters. The two dimensions, Dim1 and Dim2, are principal components derived from the original features, explaining 63% and 33.3% of the variance, respectively. We can see that the clusters are well-separated, indicating that the K-means algorithm has effectively partitioned the data into distinct groups.

#### Hierarchical Clustering

Hierarchical clustering builds a hierarchy of clusters without needing to pre-specify the number of clusters. It can be agglomerative (bottom-up), starting with individual observations and merging them, or divisive (top-down), starting with one cluster and splitting it.

##### Different linkages methods analysis

In this section, we perform hierarchical clustering on our dataset. To identify the best linkage method and the optimal number of clusters (k), we computed silhouette width values for various combinations. We consider the following linkage methods: Ward.D2, complete, average, single, and centroid linkage. For k, we test a range of values between 2 and 4, as these provided the best results in our preliminary analysis.


```{r echo=FALSE}

distance_matrix = dist(clustering_data_scaled)

methods = c("complete", "single", "average", "centroid", "ward.D2")
k_n = c(2,3,4)
results = data.frame(Method = character(), 
                     K = integer(), Silhouette_Width = numeric(), stringsAsFactors = FALSE)
par(mfrow=c(1,5))
for (method in methods){
  hc_result = hclust(distance_matrix, method = method)
  plot(hc_result, main = method)
}

```


The results of hierarchical clustering are depicted in the dendrograms. Each horizontal line represents the merging or splitting of clusters at various levels of the hierarchy, with the height of the line indicating the distance or dissimilarity between the clusters when they are joined. 
Complete Linkage offers balanced cluster compactness, while Single Linkage is prone to elongated clusters and noise sensitivity. Average Linkage produces more natural clusters, and Centroid Linkage shows size variations influenced by outliers. Ward’s Method forms compact, well-defined clusters, minimizing within-cluster variance, and is the most robust method.


```{r echo=FALSE}

# Compare silhouette at different ks with different linkages methods
for (method in methods){
  for (k in k_n){
    cluster = hclust(distance_matrix, method = method)
    silhouette_results = silhouette(cutree(cluster, k = k), distance_matrix)
    silhouette_width = mean(silhouette_results[, "sil_width"])
    results = rbind(results, data.frame(Method = method, K = k, Silhouette_Width = silhouette_width))
  }
}

kable(results, caption = "Silhouette width with different k and linkages methods")

```


The table indicates that 2 clusters provide the highest result across all clustering methods, with single linkage achieving the highest silhouette width (0.5238) for K=2. However, single linkage often results in less compact clusters. Centroid and average linkage also perform well with reasonably high silhouette widths, suggesting they may offer more compact and well-defined clusters. In summary, 2 clusters are optimal, and while single linkage performs best, centroid and average linkage methods may be preferable for more cohesive clusters.
However, since we aim to build a model that balances compactness and natural data representation Ward.D2 with k = 3 seems the best choice (in particular if we consider the dendogram) even if it has a lower silhouette width. Indeed, it creates very compact clusters and is often considered robust for many clustering tasks.


##### Hierarchical clustering in action

In order to have a more accurate representation of the segmentation, we report the dendogram cutting it at early stages so that it is more readable. 

```{r echo=FALSE}
hc_result = hclust(distance_matrix, method = "ward.D2")

# Visualize the dendrogram
plot(hc_result, cex = 0.5, main = "Dendrogram", xlab = "", sub = "", labels = FALSE)
rect.hclust(hc_result, k = 3, border = 2:5)

cutree_result_h = cutree(hc_result, k = 3)


```

Subsequently, we visualize the segments in a lower dimensional space:

```{r echo=FALSE}

# Add the cluster labels to the original data
Data$Cluster_h = as.factor(cutree_result_h)

fviz_cluster(list(data = clustering_data_scaled, cluster = cutree_result_h),
             geom = "point",
             ellipse.type = "convex",
             palette = "jco",
             ggtheme = theme_minimal(),
             main = "Cluster Visualization using Hierarchical Clustering")


```


It is clear that the clusters are similar to the ones that we have found with k-mean algorithm. However, in this case there are noticeable overlaps between different clusters, indicating that some customer segments share similar characteristics. This overlap suggests that while hierarchical clustering has identified distinct groups, the boundaries between some clusters are not entirely clear-cut, reflecting the complexity and potential similarities within the dataset. 

### CLUSTERS ANALYSIS AND CONCLUSION

#### Cluster Analysis

Since K-means represents in a clearer way the clusters without "overlapping" (problem that we encountered adopting the hierarchical method), we decided to consider the latter for our final analysis.

Initially, we analyze each cluster by plotting the mean values of the features used to construct the clusters. This allows us to understand the characteristic profiles of each cluster based on the key attributes that define them.

```{r echo=FALSE, warning=FALSE, message=FALSE}
# Analyze cluster characteristics
cluster_summary <- Data %>%
  group_by(Cluster) %>%
  summarize(across(c(Account.length, Total.minutes, CLV), mean, .names = "mean_{.col}")) %>%
  arrange(Cluster)
library(tidyr)
# Reshape the data for plotting
cluster_summary_long <- cluster_summary %>%
  pivot_longer(cols = starts_with("mean_"), names_to = "Metric", values_to = "MeanValue") %>%
  mutate(Metric = factor(Metric, levels = c("mean_Account.length", "mean_Total.minutes", "mean_CLV"),
                         labels = c("Account Length", "Total Minutes", "CLV")))

# Plotting the cluster characteristics
ggplot(cluster_summary_long, aes(x = Cluster, y = MeanValue, fill = Cluster)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ Metric, scales = "free_y") +
  theme_minimal(base_size = 15) +
  labs(
    title = "Cluster Characteristics",
    x = "Cluster",
    y = "Mean Value",
    fill = "Cluster"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.position = "none"
  )

```


The most interesting results are observed in cluster 3, where customers have the highest CLV and Total Minutes values. This cluster highlights inefficiencies that Telecom is facing: despite its high CLV, cluster 3 does not have a particularly high mean account length, which is higher in cluster 2. Generally, many clusters with high CLV exhibit low account length, suggesting that the company may be overcharging its most active consumers. An exception is cluster 2, which likely represents a well-defined segment for which the company provides high-quality services. In contrast, cluster 1 shows the lowest values across all metrics, indicating a segment with lower engagement and value. This analysis suggests the need for targeted retention strategies, especially for high CLV customers with shorter account lengths, to optimize customer satisfaction and retention.


```{r warning=FALSE, message=FALSE, echo=FALSE}

# Analyze the number of churned customers in each cluster
churn_analysis <- Data %>%
  group_by(Cluster, Churn) %>%
  summarise(Count = n()) %>%
  arrange(Cluster, Churn)

ggplot(churn_analysis, aes(x = Cluster, y = Count, fill = Churn)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal(base_size = 15) +
  labs(
    title = "Churned Customers in Each Cluster",
    x = "Cluster",
    y = "Number of Customers",
    fill = "Churn Status"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )
```


The bar plot illustrates the number of churned (True) and non-churned (False) customers across various clusters, revealing significant insights into customer retention. Overall, the three clusters are fairly equally distributed in terms of customer count. Notably, cluster 3 stands out with the highest number of churned customers, despite having the highest Customer Lifetime Value (CLV). This discrepancy suggests that Telecom is struggling to retain its most valuable customers, likely due to dissatisfaction or unmet expectations.

To address this, Telecom should implement targeted retention strategies, such as personalized offers and improved customer service, to reduce churn and maximize the value of these high-CLV customers. This analysis highlights the critical need to focus on clusters with high churn rates and high CLV for targeted retention strategies, enabling the company to optimize promotional efforts and reduce costs. By implementing focused promotional campaigns and retention initiatives specifically for cluster 3, Telecom can effectively retain its most valuable customers, enhance overall customer satisfaction, and ultimately improve profitability by reducing churn among high-value segments.


